Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-02-14 fscd19/fscd19.tex
[ fscd19 ] an introduction and a conclusion

\ccsdesc[100]{General and reference~General literature}
\ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

@@

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

@@

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.
The normal form is achieved by a type-directed \emph{reification}
procedure after evaluating the open term to a semantic value, mapping
the free variables to corresponding \emph{unknowns} in the semantics.
The literal use of a standard interpreter
can be achieved for the pure simply-typed lambda-calculus
\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
by modelling uninterpreted base types as sets of neutral (aka atomic)
terms, or more precisely, as presheaves or sets of neutral term
families, in order to facilitate fresh bound variable generation during
reification of functions to lambdas.
Thanks to $\eta$-equality at function types, free variables of
function type can be reflected into the semantics as functions
applying the variable to their reified argument,
becoming a neutral.  This mechanism provides us with unknowns of
function type which can be faithfully reified to normal forms.
% Since neutrals are closed under application to normal arguments, they
% can act as unknown functions, storing the reified semantic argument
% they are invoked with (thus, these unknowns are aka accumulators
% \cite{gregoireleroy:icfp02}).

For the extension to sum types (logically disjunctions and categorically
weak coproducts), this reflection trick does not work anymore.  A
semantic value of binary sum type is either a left or a right
injection, but the decision between left or right cannot be taken at
reflection time, since a variable of sum type does not provide us with
the information.  A literal standard interpreter for closed terms can
thus no longer be used for NbE.  Instead, we can utilize a monadic
interpreter.  When the interpreter attempts a case distinction on an
unknown of sum type, it asks an oracle whether the unknown is a left
or a right injection.  The oracle returns one of these alternatives,
wrapping a new unknown in the injection.  The communication with the
oracle can be modeled in a \emph{monad} $\C$,
% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
which records the questions asked and the continuation of the
interpreter for each of the possible answers.
A monadic semantic value is thus a \emph{case tree} where the leaves
are occupied by non-monadic values.
In this article we only consider weak sum types, producing non-unique
normal forms, where it does not
matter in which order the questions are asked (commuting case splits),
and whether the same
question is asked several times (redundant case splits).
The model would need refinement for strong, extensional sums
\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.

Filinski \cite{filinski:tlca01} studied NbE for Moggi's computional lambda
calculus \cite{moggi:infcomp91} shedding light on the difference
between call-by-name (CBN) and call-by-value (CBV) NbE, where Danvy's
type-directed partial evaluation \cite{danvy:popl96} falls into the
latter class.  The contribution of the computational lambda calculus
is to make explicit where the monad is envoked during monadic
evaluation, and this placement of the monad carries over to the NbE
setting.  Moggi's studies were continued by Levy \cite{levy:hosc06}
who designed the call-by-push-value (CBPV) lambda-calculus to embed
both the CBN and CBV lambda calculus.

@@

In contrast to the normal forms of CBN NbE, which is the algorithmic
counterpart of the completeness proof for intuitionistic propositional
logic (IPL) using Beth models, CBPV NbE gives us more restrained
normal forms, where the production of a value via injections cannot be
interrupted by more questions to the oracle.  In the research field of
focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
\emph{chaining non-invertible introductions}.
Invertible introductions are already chained in NbE thanks to
extensionality ($\eta$) for function (an more generally, negative)
types.  Non-invertible eliminations are also happing in a chain,
building neutrals.  What is missing from the picture is the chaining
of invertible eliminations, i.e., case distinctions and, more
generally, pattern matching.  The picture is completed by extending
NbE to \emph{polarized lambda calculus}
\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
in Section~\ref{sec:fipl}.

In our presentation of the various lambda calculi and NbE, we use
presheaves for the abstract syntax and the semantics
\cite{altenkirchHofmannStreicher:ctcs95}.

In contrast to Zeilberger \cite{zeilberger:PhD}.

@@

>>>>

% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10003752.10003790.10011740</concept_id>
% <concept_desc>Theory of computation~Type theory</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% <concept>
% <concept_id>10003752.10010124.10010125.10010127</concept_id>
% <concept_desc>Theory of computation~Functional constructs</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% <concept>
% <concept_id>10003752.10010124.10010125.10010130</concept_id>
% <concept_desc>Theory of computation~Type structures</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% <concept>
% <concept_id>10003752.10010124.10010131.10010137</concept_id>
% <concept_desc>Theory of computation~Categorical semantics</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% <concept>
% <concept_id>10003752.10003790.10003792</concept_id>
% <concept_desc>Theory of computation~Proof theory</concept_desc>
% <concept_significance>100</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}

\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Theory of computation~Type structures}
\ccsdesc[300]{Theory of computation~Functional constructs}
\ccsdesc[300]{Theory of computation~Proof theory}
\ccsdesc[300]{Theory of computation~Categorical semantics}
\ccsdesc[100]{Theory of computation~Operational semantics}
%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

@@

\supplement{\url{https://andreasabel.github.io/ipl/html/NfModelMonad.html}}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

@@

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.
The normal form is achieved by a type-directed \emph{reification}
procedure after evaluating the open term to a semantic value, mapping
(\emph{reflecting})
the free variables to corresponding \emph{unknowns} in the semantics.
The literal use of a standard interpreter
can be achieved for the pure simply-typed lambda-calculus
\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
by modelling uninterpreted base types as sets of neutral (aka atomic)
terms, or more precisely, as presheaves or sets of neutral term
families, in order to facilitate fresh bound variable generation during
reification of functions to lambdas.
Thanks to $\eta$-equality at function types, free variables of
function type can be reflected into the semantics as functions
applying the variable to their reified argument,
forming a neutral term.  This mechanism provides us with unknowns of
function type which can be faithfully reified to normal forms.
% Since neutrals are closed under application to normal arguments, they
% can act as unknown functions, storing the reified semantic argument
% they are invoked with (thus, these unknowns are aka accumulators
% \cite{gregoireleroy:icfp02}).

For the extension to sum types (logically, disjunctions and categorically,
weak coproducts), this reflection trick does not work anymore.  A
semantic value of binary sum type is either a left or a right
injection, but the decision between left or right cannot be taken at
reflection time, since a variable of sum type does not provide us with
the information.  A literal standard interpreter for closed terms can
thus no longer be used for NbE;  instead, we can utilize a monadic
interpreter.  When the interpreter attempts a case distinction on an
unknown of sum type, it asks an oracle whether the unknown is a left
or a right injection.  The oracle returns one of these alternatives,
wrapping a new unknown in the injection.  The communication with the
oracle can be modeled in a \emph{monad} $\C$,
% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
which records the questions asked and the continuation of the
interpreter for each of the possible answers.
A monadic semantic value is thus a \emph{case tree} where the leaves
are occupied by non-monadic values \cite{altenkirchUustalu:flops04}.
In this article we only consider weak sum types, producing non-unique
normal forms, where it does not
matter in which order the questions are asked (commuting case splits),
and whether the same
question is asked several times (redundant case splits).
The model would need refinement for strong, extensional sums
\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.

Filinski \cite{filinski:tlca01} studied NbE for Moggi's computional lambda
calculus \cite{moggi:infcomp91}, shedding light on the difference
between call-by-name (CBN) and call-by-value (CBV) NbE, where Danvy's
type-directed partial evaluation \cite{danvy:popl96} falls into the
latter class.  The contribution of the computational lambda calculus
is to make explicit where the monad is invoked during monadic
evaluation, and this placement of the monad carries over to the NbE
setting.  Moggi's studies were continued by Levy \cite{levy:hosc06}
who designed the call-by-push-value (CBPV) lambda-calculus to embed
both the CBN and CBV lambda calculus.

@@

In contrast to the normal forms of CBN NbE, which is the algorithmic
counterpart of the completeness proof for intuitionistic propositional
logic (IPL) using Beth models, CBPV NbE gives us more restrained
normal forms, where the production of a value via injections cannot be
interrupted by more questions to the oracle.  In the research field of
focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
\emph{chaining non-invertible introductions}.
Invertible introductions are already chained in NbE thanks to
extensionality ($\eta$) for function, and more generally, negative
types.  Non-invertible eliminations are also happening in a chain,
building neutrals.  What is missing from the picture is the chaining
of invertible eliminations, i.e., case distinctions and, more
generally, pattern matching.  The picture is completed by extending
NbE to \emph{polarized lambda calculus}
\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
in Section~\ref{sec:fipl}.

In our presentation of the various lambda calculi we ignore the
concrete syntax, only consider the abstract syntax obtained by the
Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
nodes are rule invokations.  Thus, a intrinsically typed,
nameless syntax is most natural,
and our syntactic classes are just presheaves over the category of
typing contexts and renamings.  The use of presheaves then smoothly
extents to the semantic constructions \cite{altenkirchHofmannStreicher:ctcs95}.

Concerning the presentation of polarized lambda calculus, we depart from
Zeilberger \cite{zeilberger:PhD} who employs \emph{a priori}
infinitary syntax, modelling a case tree as a meta-level function mapping
well-typed patterns to branches.  Instead, we use a graded monad
representing complete pattern matching over a newly added hypothesis,
which is in spirit akin to Filinski's \cite[Section~4]{filinski:tlca01}
and Krishnaswami's \cite{krishnaswami:popl09} treatment of eager
pattern matching using a separate context of variables to be matched
on.

Our design choices were guided by an Agda formalization of sections 2
(complete) and 4 (partial), available at \url{https://github.com/andreasabel/ipl}.
Agda was particularly helpful to correctly handle the renamings
abundantly present when working with presheaves.

@@

We have implemented NbE for CBPV and polarized lambda calculus
formulated with intrinsically well-typed syntax and presheaf
semantics.  As a side result, we have proven semantically
that the normal forms of both systems are logically complete, \ie,
each derivable judgement $\Gamma \der N$ has a normal derivation.
It remains to show that NbE for these calculi is also computationally
sound and complete, i.e., the computational behavior of term and
normal form should agree, and normalization should decide a suitable
equational theory on terms.

Additionally, a natural question to investigate is whether known CBN
and CBV NbE algorithms can be obtained from our NbE algorithms by
embedding simply-typed lambda calculus into our polarized calculi,
using known CBN and CBV translations.  Further, we would like to study
the NbE algorithm for STLC arising from the optimal translation, i.e.,
the one inserting a minimal amount of $\tThunk$ and $\tComp$ transitions.
